---
title: "Water Footprint Analysis"
author: "Stephen Franklin"
date: "May 18, 2015"
output: html_document
---

# Water Footprint Differences Between Animal-Sourced and Plant-Sourced Foods.

In this study we quantify the differences in water efficiency (cubic meters per metric ton) between foods of animal and plant origin. We found no significant difference between foods of the two origins, but we did find that a better indicator for water footprint efficiency is the proportion of protein in the food. Furthermore, meats and nuts have both the highest amounts of protein as well as extremely high water footprints. The best model predicts that for each 1 percent protein in a food, it would take ___ cubic meters per ton of water to produce. That model explains 83% of the variance in the data.

```{r libraries, echo=FALSE, message=F}
require(knitr)
library(data.table)
library(ggplot2)
library(jsonlite)
library(curl)
library(plyr)
library(rCharts)
library(psych)
library(xtable)
#library(printr)
load("water_footprint_table.RData")
```

```{r}
setkey(water,protein)
rm(California_footprint, Global_avg_footprint, Products, protein)
attach(water)
```

```{r}
require(psych)
par(mfrow = c(1,1))
pairs.panels( water[,c(2:8),with=F], smooth=T, ellipses=F, cor=T,
      main = "Water Footprint Correlations", pch=21, 
      bg = 2 + (as.numeric(water$kingdom) < 2)) ## color: red=2,green=3, plant=1,animal=2
```

Looking at the pairs plot, it seems there's no clear distinction between animal-sourced foods and plant-sourced foods based upon these basic nutritional components. There are some interesting things. California seems to have a few foods that have very different water footprints compared to the global average. Protein seems to correlate positively with water footprint, as does energy. Fat somewhat correlates with water footprint globally, but not at all in California. Carbohydrates correlate not at all, and water content correlates negatively with water footprint.

Also, in many of the mini-plots the data seems to clump together in one corner and disperse, increasing in variance, across the diagonal.

### Overall Model

```{r}
wfit0 <- lm(Global_avg_footprint ~ protein + fat + water + carb + kcal + kingdom, water)
summary(wfit0)$r.squared
summary(wfit0)$coefficients
```

In this overall weighted model, not many of these variables seem significant, and the model only explains 69% of the variance. The only variable with a marginally significant p-value is protein at **0.054**.

Let's look at water footprint versus protein alone.
```{r}
wfit1 <- lm(Global_avg_footprint ~ protein, water)
summary(wfit1)$r.squared
summary(wfit1)$coefficients
```
Protein alone explains **56%** of the variance, but with a strong p-value.
The intercept itself, however, is not significant.

So based on an unpredictable intercept, we would expect a 360 m3/ton increase in the water footprint of a food for each 1% increase in its typical measure of protein.

### Do animal-based foods have a bigger water footprint than plant-based foods?

```{r}
wfit2 <- lm(Global_avg_footprint ~ as.numeric(kingdom), water)
summary(wfit2)$r.squared
summary(wfit2)$coefficients
```
It seems the answer is no.


Let's look at energy:
```{r}
wfit3 <- lm(Global_avg_footprint ~ kcal, water)
summary(wfit3)$r.squared
summary(wfit3)$coefficients
```
Not surprisingly, energy is positively correlated with water footprint. We would expect (upon a still unpredictable intercept) a 15 m3/ton increase in water footprint for each kilocalorie increase in a food.

Just for fun, let's look for an interaction between kingdom and protein
```{r}
wfit4 <- lm(Global_avg_footprint ~ protein * kingdom, water)
summary(wfit4)$r.squared
summary(wfit4)$coefficients
```

The reference category for `kingdom` is "1" -- plants, 
which has an (insignificant) intercept of 218 m3/ton. 
The slope for *animal* is 396. 
The change in intercept for *animal* is 1189.
And the change in slope for *animal* is -126.

The interaction isn't significant (p: 0.26 > 0.05)

Let's plot the regression lines for animal versus plant sources,
with protein as an interaction:
```{r}
par(mfrow = c(1,1))
plot(protein,Global_avg_footprint,pch=19)
points(protein,Global_avg_footprint,pch=19,col=((as.numeric(kingdom)<2)*1+2))

## Again, the as.numeric() turns the factor "0" into a 1
## and the factor "1" into a 2.

abline(c(wfit4$coeff[1],wfit4$coeff[2]),col="green",lwd=3)
abline(c(wfit4$coeff[1] + wfit4$coeff[3],wfit4$coeff[2] +wfit4$coeff[4]),col="red",lwd=3)
```

The two food sources seem closely correlated, and that agrees with the interaction being insignificant.

But let's check the residuals:
```{r}
par(mfrow = c(1,1))
ew <- wfit4$residuals 
plot(kingdom, ew,
     xlab = "kingdom (0=plant)",
     ylab = "Residuals")
```

compared to fit1 (water footprint vs. protein):
```{r}
par(mfrow = c(1,1))
ew <- wfit1$residuals 
plot(kingdom, ew,
     xlab = "kingdom (0=plant)",
     ylab = "Residuals")
```
Very similar. Lots of outliers in plants.

Resdiuals and Leverage for water footprint vs. protein:
```{r}
par(mfrow = c(2, 2))
plot(wfit1)
```

Three products -- cashew, beef, and almond -- have somewhat more leverage in the model. But their leverage is less than 0.5, so they don't bias the model very much. 

Analysis of variance to compare the models:
```{r}
anova(wfit0, wfit1, wfit2, wfit3, wfit4)
```

Somewhat strangley, anova shows that wfit4 (predicting the footprint by kingdom holding protein constant) is the best model.

Let's see what an automatic fit does:
```{r}
library(MASS)
wfit.s <- stepAIC(lm(Global_avg_footprint ~ ., data = water[,c(3:8),with=F]), trace = 0) 
summary(wfit.s)$coeff
summary(wfit.s)$r.squared
```

Interestingly, the stepAIC found a model that explains 69% of the variance by including water as well as protein and energy.

Still, protein has the highest significance. So for simplicity, I'd like to focus on just that one factor. I'll show kingdom because I think it's non-significance is interesting and relevant.

```{r}
w1 <- ggplot(data=water, aes(y = Global_avg_footprint, x = protein, color=kingdom)) +
        geom_point(data = subset(water, protein <= 3 & Global_avg_footprint < 3000))
w1 <- w1 + scale_color_manual(breaks = c("plant", "animal"),
                              values=c("darkred", "green4"),
                              name= "Source")
w1 <- w1 + geom_smooth(data = water, method="lm", fullrange=TRUE)
w1 <- w1 + geom_text(aes(label=ifelse(
                                protein > 3 |
                                Global_avg_footprint > 3000,
                                Products,'')),
                     hjust=0,just=0, size=3,
                     position = position_jitter(w = 0.5, h = 300),
                     show_guide = F)
w1 <- w1 + labs(y=expression("Water Footprint: Global Average ("*m^3/ton*")"),
            x = "Protein (%)", title="Water Footprint vs. Protein of Raw Foods: Global Average")
w1 <- w1 + theme(axis.title=element_text(face="bold.italic",
   color="brown"), legend.position="top")
w1 <- w1 + expand_limits(y = c(0, 17500), x = c(0, 30) )
w1
```

